---
title: "chapter2_trees_results"
author: "Leticia Magpali"
format: html
editor: visual
---

## Chapter 2 - Phylogenetic trees

### 1. Importing packages

```{r}
library(ape)
library(tidyverse)
library(ggplot2)
library(ggtree)
```

### 2. Loading tree files

ðŸŒ³ **McGowen et al. (2019) species tree**

```{r}
McGowen2019_tree <- read.tree("trees/species_trees/RAxML_partition_by_gene_best_tree.tre")
```

*2.1. Small data set*

ðŸ„ Species tree

```{r}
species_tree_smalldata <- read.tree("trees/species_trees/small_dataset/small-dataset-tree.phy")
```

ðŸ„ Codon trees

```{r}

ACTC1_tree_sc <- read.tree("trees/gene_trees/small_dataset/ACTC1_tree/ACTC1_codon.treefile")

CALML6_tree_sc <- read.tree("trees/gene_trees/small_dataset/CALML6_tree/CALML6_codon.treefile")

CASQ1_tree_sc <- read.tree("trees/gene_trees/small_dataset/CASQ1_tree/CASQ1_codon.treefile")

CASQ2_tree_sc <- read.tree("trees/gene_trees/small_dataset/CASQ2_tree/CASQ2_codon.treefile")

CNTNAP2_tree_sc <- read.tree("trees/gene_trees/small_dataset/CNTNAP2_tree/CNTNAP2_codon.treefile")

FOXP2_tree_sc <- read.tree("trees/gene_trees/small_dataset/FOXP2_tree/FOXP2_codon.treefile")

HSP90AA1_tree_sc <- read.tree("trees/gene_trees/small_dataset/HSP90AA1_tree/HSP90AA1_codon.treefile")

HTT_tree_sc <- read.tree("trees/gene_trees/small_dataset/HTT_tree/HTT_codon.treefile")

MYH2_tree_sc <- read.tree("trees/gene_trees/small_dataset/MYH2_tree/MYH2_codon.treefile")

MYL1_tree_sc <- read.tree("trees/gene_trees/small_dataset/MYL1_tree/MYL1_codon.treefile")

MYO7A_tree_sc <- read.tree("trees/gene_trees/small_dataset/MYO7A_tree/MYO7A_codon.treefile")

NLGN3_tree_sc <- read.tree("trees/gene_trees/small_dataset/NLGN3_tree/NLGN3_codon.treefile")

NRXN1_tree_sc <- read.tree("trees/gene_trees/small_dataset/NRXN1_tree/NRXN1_codon.treefile")

OTOF_tree_sc <- read.tree("trees/gene_trees/small_dataset/OTOF_tree/OTOF_codon.treefile")

PCDH15_tree_sc <- read.tree("trees/gene_trees/small_dataset/PCDH15_tree/PCDH15_codon.treefile")

RYR2_tree_sc <- read.tree("trees/gene_trees/small_dataset/RYR2_tree/RYR2_codon.treefile")
```

*2.2. Large data set*

### 3. Visualizing trees with ggplot

*3.1. Species trees*

*ðŸ„ McGowen et al.*

```{r}
ggtree(McGowen2019_tree) + geom_treescale() + geom_tiplab()
```

*3.2. Codon trees, small data set*

Trees are visualized and checked for: longest/shortest branches, mismatches with species tree

ðŸ„ ACTC1

-   Phocenidae (Psin+Naasi) looses monophyly

```{r}
ggtree(ACTC1_tree_sc) + geom_treescale() + geom_tiplab()
```

ðŸ„ CALML6

-   Phocenidae (Psin+Naasi) looses monophyly

```{r}
ggtree(CALML6_tree_sc) + geom_treescale() + geom_tiplab()
```

ðŸ„ CASQ1

-   Igeo + Pbla broken

-   C.heavisidi, C.commersonii, L.australis broken by Lbor

```{r}
ggtree(CASQ1_tree_sc) + geom_treescale() + geom_tiplab()
```

ðŸ„ CASQ2

-   Igeo + Pbla monophyly broken

```{r}
ggtree(CASQ2_tree_sc) + geom_treescale() + geom_tiplab()
```

ðŸ„ CNTNAP2

-   Igeo + Pbla monophyly broken

-   C.heavisidi, C.commersonii, L.australis broken by Lbor

```{r}
ggtree(CNTNAP2_tree_sc) + geom_treescale() + geom_tiplab()
```

ðŸ„ FOXP2

-   O.brevirostris + O.heinsohni broken

-   C.heavisidi, C.commersonii, L.australis broken

```{r}
ggtree(FOXP2_tree_sc) + geom_treescale() + geom_tiplab()
```

ðŸ„ HSP90AA1

-   O.brevirostris + O.heinsohni broken

-   Igeo + Pbla monophyly broken

```{r}
ggtree(HSP90AA1_tree_sc) + geom_treescale() + geom_tiplab()
```

ðŸ„ HTT

-   All foreground clades preserved!

```{r}
ggtree(HTT_tree_sc) + geom_treescale() + geom_tiplab()
```

ðŸ„ MYH2

```{r}
ggtree(MYH2_tree_sc) + geom_treescale() + geom_tiplab()
```

ðŸ„ MYL1

-   Phocenidae broken

```{r}
ggtree(MYL1_tree_sc) + geom_treescale() + geom_tiplab()
```

ðŸ„ MYO7A

-   All foreground clades preserved!

```{r}
ggtree(MYO7A_tree_sc) + geom_treescale() + geom_tiplab()
```

ðŸ„ NLGN3

-   All foreground clades preserved! (we have 5 of them)

```{r}
ggtree(NLGN3_tree_sc) + geom_treescale() + geom_tiplab()
```

ðŸ„ NRXN1

-   All foreground clades preserved! (but only 3 of them here)

```{r}
ggtree(NRXN1_tree_sc) + geom_treescale() + geom_tiplab()
```

ðŸ„ OTOF

-   C.heavisidi, C.commersonii, L.australis broken by Lbor

```{r}
ggtree(OTOF_tree_sc) + geom_treescale() + geom_tiplab()
```

ðŸ„ PCDH15

-   All foreground clades preserved!

```{r}
ggtree(PCDH15_tree_sc) + geom_treescale() + geom_tiplab()
```

ðŸ„ RYR2

-   All foreground clades preserved!

```{r}
ggtree(RYR2_tree_sc) + geom_treescale() + geom_tiplab()
```
