## Installing packages

```{r}
library(ggplot2)
library(dplyr)
library(MASS)
library(mvnormtest)
library(tidyr)
```

# Analyses with non-transformed values #########
# ##############################################

## Observing and exploring dataset

```{r}

# Loading in data
echo <- read.csv("acoustic_features/acoustic_features - Echo.csv")
# echo1 <- read.csv("Acoustic parameters - Selection1.csv")
echo2 <- read.csv("acoustic_features/acoustic_features - echo_selection2.csv")
# echo3 <- read.csv("Acoustic parameters - Selection3.csv")

summary(echo)

attach(echo)

# Exploratory plot to identify potentially colinear variables
plot(echo)

# Convert to numeric
echo$ICI <- as.numeric(echo$ICI)

# Log transformation
echo_numeric_columns <- data.frame(echo[3:14])
echo_log_columns <- log(echo_numeric_columns)
echo_log <- cbind(echo[1:2], echo_log_columns)

```

## Exploratory plots

```{r}

environment <- c("turquoise3", "purple", "seagreen4", "salmon2","orangered2")[as.factor(Environment)]

# Centroid frequency by peak frequency
plot(Fc~Fp, col = alpha(environment, 0.6),
     cex = 1.5, pch = 19, xlab = "Peak frequency", ylab = "Centroid Frequency", 
     main = "Scatterplot of echolocation frequency")

legend("bottomright", 1, pch = 19, legend = c("Riverine", "Coastal","Oceanic", "Polar", "Mixed")[unique(as.numeric(factor(Environment)))], col = c("turquoise3", "purple", "seagreen4", "salmon2","orangered2")[unique(as.numeric(factor(Environment)))])



# Centroid frequency by SLRMS
plot(echo$Fc~echo$SLRMS, col = alpha(environment, 0.6),
     cex = 1.5, pch = 19, xlab = "Peak frequency", ylab = "Centroid Frequency")



# Centroid frequency by peak to peak source level
plot(echo$Fc~echo$SLpp,  col = alpha(environment, 0.6),
     cex = 1.5, pch = 19, xlab = "Centroid frequency", ylab = "Peak to peak source level")



# Peak to peak source by SLRMS
plot(echo$SLpp~echo$SLRMS, col = alpha(environment, 0.6),
     cex = 1.5, pch = 19, xlab = "RMS source level", ylab = "Peak to peak source level")


# Box plotting variables by environment
boxplot(Fc~Environment, col = "lavender", border = "purple", ylab = "Centroid Frequency")
boxplot(Fp~Environment, col = "lavender", border = "purple", ylab = "Peak Frequency")
boxplot(SLpp~Environment, col = "lavender", border = "purple", ylab = "Peak to peak source levels")
boxplot(SLEFD~Environment)
boxplot(SLRMS~Environment)
boxplot(Duration~Environment)
boxplot(ICI~Environment)
boxplot(BWRMS~Environment)

```

## MANOVA

```{r}

# Source levels
sl <- echo[, 1:5] %>% drop_na()
mano_echo <- manova(cbind(SLpp, SLRMS, SLEFD) ~ Environment, data = sl)
summary(mano_echo, test = "Wilks")

# Frequency
freq <- dplyr::select(echo, c(Species, Environment, Fc, Fp)) %>% drop_na()
mano_echo2 <- manova(cbind(Fc, Fp) ~ Environment, data = freq)
summary(mano_echo2, test = "Wilks")

# Frequency + Source levels peak to peak
freq_sl <- dplyr::select(echo, c(Species, SLpp, Environment, Fc, Fp)) %>% drop_na()
mano_echo3 <- manova(cbind(SLpp, Fc, Fp) ~ Environment, data = freq_sl)
summary(mano_echo3, test = "Wilks") 

# Bandwidth
band <- dplyr::select(echo, c(Species, Environment, BW3dB, BWRMS, QRMS)) %>% drop_na()
mano_echo4 <- manova(cbind(BW3dB, BWRMS, QRMS) ~ Environment, data = band)
summary(mano_echo4, test = "Wilks") 

# Duration/ICI
dur <- dplyr::select(echo, c(Species, Environment, Duration, ICI)) %>% drop_na()
mano_echo5 <- manova(cbind(Duration, ICI) ~ Environment, data = dur)
summary(mano_echo5, test = "Wilks")

# All factors combined:
echo_all <- dplyr::select(echo, c(Species, Environment, SLpp, Fc, Fp, BWRMS)) %>% drop_na()
mano_echo6 <- manova(cbind(SLpp, Fc, Fp, BWRMS) ~ Environment, data = echo_all)
summary(mano_echo6, test = "Wilks")


# Checking MANOVA assumptions

# Normality
mshapiro.test(t(mano_echo$residuals))
mshapiro.test(t(mano_echo2$residuals)) 
mshapiro.test(t(mano_echo3$residuals)) 
mshapiro.test(t(mano_echo4$residuals))
mshapiro.test(t(mano_echo5$residuals))
mshapiro.test(t(mano_echo6$residuals))

# Remove NHBF species?? Because they are a different case
```

## LDA or Canonical Variate Analysis -- model 1

```{r}

attach(freq_sl)

# Setting up color schemes
environment_lda <- c("turquoise3", "purple", "goldenrod1", "indianred3","lightsalmon")[as.factor(Environment)]

lda_echo <- lda(Environment ~ Fc + Fp + SLpp, data = freq_sl)
summary(lda_echo)

# Calculating discriminant scores:

lscore_echo <- cbind(Fc, Fp, SLpp)%*%lda_echo$scaling 

#Plotting each measurement score against the Species variable 
plot(x = lscore_echo[,1],lscore_echo[,2],
     col = alpha(environment_lda, 0.5),
     pch = 19, 
     cex = 1.5,
     xlab = "1st canonical variate",
     ylab = "2nd canonical variate",
     asp = 1,
     main = "Discriminant analysis")

legend("topright", 1, pch = 19, 
       legend = c("Riverine", "Coastal","Oceanic", "Polar", "Mixed")
       [unique(as.factor(Environment))], col = unique(environment_lda))

#Plotting with ggplot - for later
# ggplot(lscore_echo, aes(x = lscore_echo[,1], y = lscore_echo[,2])) +
#   geom_point()

```


## LDA or Canonical Variate Analysis -- model 2

```{r}

attach(echo_all)

# Setting up color schemes
environment_lda2 <- c("turquoise3", "purple", "goldenrod1", "indianred3","lightsalmon")[as.factor(Environment)]

lda_echo2 <- lda(Environment ~ SLpp + Fc + Fp + BWRMS, data = echo_all)
lda_echo2

# Calculating discriminant scores:

lscore_echo2 <- cbind(Fc, Fp, SLpp, BWRMS)%*%lda_echo2$scaling 

#Plotting each measurement score against the Species variable 
plot(x = lscore_echo2[,1],lscore_echo2[,2],
     col = alpha(environment_lda2, 0.5),
     pch = 19, 
     cex = 1.5,
     xlab = "1st canonical variate",
     ylab = "2nd canonical variate",
     asp = 1,
     main = "Discriminant analysis")

legend("topleft", 1, pch = 19, legend = c("Riverine", "Coastal","Oceanic", "Polar", "Mixed")
       [unique(as.factor(Environment))], col = unique(environment_lda2))

```


## PCA analysis  

```{r}

attach(echo2)

environment_pca <- c("turquoise3", "purple", "goldenrod1", "indianred3","lightsalmon")[as.factor(Environment)]

echo_PCA <- princomp(echo2[, 3:11], cor = TRUE, scores = TRUE)
summary(echo_PCA)

biplot(echo_PCA)

plot(echo_PCA$scores[,1], echo_PCA$scores[,2],
     col = alpha(environment_pca, 0.6),
     cex = 1.5, pch = 19)

plot(echo_PCA$scores[,2], echo_PCA$scores[,3],
     col = alpha(environment_pca, 0.6),
     cex = 1.5, pch = 19)

#Varimax rotation for PCs 1 and 2
echo.var <- varimax(echo_PCA$loadings[,1:2])
echo.vscores <- echo_PCA$scores[,1:2]%*%echo.var$rotmat  
echo.vscores.df <- data.frame(echo.vscores[,1:2])

#Plotting varimax scores 
plot(echo.vscores[,1], echo.vscores[,2],
     pch = 19,
     asp = 1,
     cex = 1.5,
     xlab = "1st varimax component",
     ylab = "2nd varimax component", 
     main = "Varimax scores plot",
     col = alpha(environment_pca, 0.6))

legend("topleft", 1, pch = 19, legend = c("Riverine", "Coastal","Oceanic", "Polar", "Mixed")
       [unique(as.factor(Environment))], col = unique(environment_pca))

#Adding loadings and names of the variables to each arrow that shows the loading
arrows(4,0,echo.var$loadings[,1],echo.var$loadings[,2],col="grey", length = 0.1)
text(echo.var$loadings[,1], echo.var$loadings[,2], names(echo2[, 3:11]), cex = 0.6, col = "grey")



#Varimax rotation for PCs 2 and 3
echo.var2 <- varimax(echo_PCA$loadings[,2:3])
echo.vscores2 <- echo_PCA$scores[,2:3]%*%echo.var$rotmat  
echo.vscores.df2 <- data.frame(echo.vscores[,1:2]) #this is for ggplot

#Plotting varimax scores 
plot(echo.vscores2[,1], echo.vscores2[,2],
     pch = 19,
     asp = 1,
     cex = 1.5,
     xlab = "1st varimax component",
     ylab = "2nd varimax component", 
     main = "Varimax scores plot",
     col = alpha(environment_pca, 0.6))

legend("topright", 1, pch = 19, legend = c("Riverine", "Coastal","Oceanic", "Polar", "Mixed")
       [unique(as.factor(environment_pca))], col = unique(environment_pca))

#Adding loadings and names of the variables to each arrow that shows the loading
arrows(2,1,echo.var2$loadings[,1],echo.var2$loadings[,2],col="black", length = 0.1)



# Visualizing results with ggplot:

ggplot(echo.vscores.df, aes(X1, X2)) +
  geom_point(aes(color = environment_pca)) +
  xlab("1st varimax component") + ylab("2nd varimax component")

```

# Analyses with log-transformed values #########
# ##############################################

```{r}
attach(echo_log)
```

## Exploratory plots

```{r}

environment <- c("turquoise3", "purple", "seagreen4", "salmon2","orangered2")[as.factor(Environment)]

# Centroid frequency by peak frequency
plot(Fc~Fp, col = alpha(environment, 0.6),
     cex = 1.5, pch = 19, xlab = "Peak frequency", ylab = "Centroid Frequency", 
     main = "Scatterplot of echolocation frequency")

legend("bottomright", 1, pch = 19, legend = c("Riverine", "Coastal","Oceanic", "Polar", "Mixed")[unique(as.numeric(factor(Environment)))], col = c("turquoise3", "purple", "seagreen4", "salmon2","orangered2")[unique(as.numeric(factor(Environment)))])



# Centroid frequency by SLRMS
plot(echo$Fc~echo$SLRMS, col = alpha(environment, 0.6),
     cex = 1.5, pch = 19, xlab = "Peak frequency", ylab = "Centroid Frequency")



# Centroid frequency by peak to peak source level
plot(echo$Fc~echo$SLpp,  col = alpha(environment, 0.6),
     cex = 1.5, pch = 19, xlab = "Centroid frequency", ylab = "Peak to peak source level")



# Peak to peak source by SLRMS
plot(echo$SLpp~echo$SLRMS, col = alpha(environment, 0.6),
     cex = 1.5, pch = 19, xlab = "RMS source level", ylab = "Peak to peak source level")


# Box plotting variables by environment
boxplot(Fc~Environment, col = "lavender", border = "purple", ylab = "Centroid Frequency")
boxplot(Fp~Environment, col = "lavender", border = "purple", ylab = "Peak Frequency")
boxplot(SLpp~Environment, col = "lavender", border = "purple", ylab = "Peak to peak source levels")
boxplot(SLEFD~Environment)
boxplot(SLRMS~Environment)
boxplot(Duration~Environment)
boxplot(ICI~Environment)
boxplot(BWRMS~Environment)

```

## MANOVA

```{r}

# Source levels
sl_log <- echo_log[, 1:5] %>% drop_na()
mano_echolog_sl <- manova(cbind(SLpp, SLRMS, SLEFD) ~ Environment, data = sl_log)
summary(mano_echolog_sl, test = "Wilks")

# Frequency
freq_log <- dplyr::select(echo_log, c(Species, Environment, Fc, Fp)) %>% drop_na()
mano_echolog_freq <- manova(cbind(Fc, Fp) ~ Environment, data = freq_log)
summary(mano_echolog_sl, test = "Wilks")

# Frequency + Source levels peak to peak
freq_sl_log <- dplyr::select(echo_log, c(Species, Environment, SLpp, SLRMS, SLEFD, Fc, Fp)) %>% drop_na()
mano_echolog_freq_sl <- manova(cbind(SLpp, Fc, Fp) ~ Environment, data = freq_sl_log)
summary(mano_echolog_freq_sl, test = "Wilks") 

# Bandwidth
band_log <- dplyr::select(echo_log, c(Species, Environment, BW3dB, BWRMS, QRMS)) %>% drop_na()
mano_echolog_band <- manova(cbind(BW3dB, BWRMS, QRMS) ~ Environment, data = band_log)
summary(mano_echolog_band, test = "Wilks") 

# Duration/ICI
dur_log <- dplyr::select(echo_log, c(Species, Environment, Duration, ICI)) %>% drop_na()
mano_echolog_dur <- manova(cbind(Duration, ICI) ~ Environment, data = dur_log)
summary(mano_echolog_dur, test = "Wilks")

# All factors combined with most species:
echo_log_all1 <- dplyr::select(echo_log, c(Species, Environment, SLpp, SLRMS, Fc, Fp, BW3dB, Duration)) %>% drop_na()
mano_echolog_all1 <- manova(cbind(SLpp, SLRMS, Fc, Fp, BW3dB, Duration) ~ Environment, data = echo_log_all1)
summary(mano_echolog_all1, test = "Wilks")

# All factors combined:
echo_log_all <- dplyr::select(echo_log, c(Species, Environment, SLpp, SLRMS, Fc, Fp, BWRMS, BW3dB, QRMS, Duration, ICI)) %>% drop_na()
mano_echolog_all <- manova(cbind(SLpp, SLRMS, Fc, Fp, BWRMS, BW3dB, QRMS, Duration, ICI) ~ Environment, data = echo_log_all)
summary(mano_echolog_all, test = "Wilks")


# Checking MANOVA assumptions

# Normality
mshapiro.test(t(mano_echolog_sl$residuals))
mshapiro.test(t(mano_echolog_freq$residuals)) 
mshapiro.test(t(mano_echolog_freq_sl$residuals)) 
mshapiro.test(t(mano_echolog_band$residuals))
mshapiro.test(t(mano_echolog_dur$residuals))
mshapiro.test(t(mano_echolog_all1$residuals))
```

## LDA or Canonical Variate Analysis -- model 1 (frequency and peak to peak source values)

```{r}

attach(freq_sl_log)

# Setting up color schemes
environment_lda_log <- c("turquoise3", "purple", "seagreen4", "salmon2","orangered2")[as.factor(Environment)]

lda_echo_log <- lda(Environment ~ Fc + Fp + SLpp, data = freq_sl_log)
summary(lda_echo_log)

# Calculating discriminant scores:

lscore_echo_log <- cbind(Fc, Fp, SLpp)%*%lda_echo_log$scaling 

#Plotting each measurement score against the Species variable 
plot(x = lscore_echo_log[,1],lscore_echo_log[,2],
     col = alpha(environment_lda_log, 0.5),
     pch = 19, 
     cex = 1.5,
     xlab = "1st canonical variate",
     ylab = "2nd canonical variate",
     asp = 1,
     main = "Discriminant analysis")

legend("topright", 1, pch = 19, 
       legend = c("Riverine", "Coastal","Oceanic", "Polar", "Mixed")
       [unique(as.factor(Environment))], col = unique(environment_lda_log))

```

## LDA or Canonical Variate Analysis -- model 2 (frequency)

```{r}

attach(freq_log)

# Setting up color schemes
environment_lda_log2 <- c("turquoise3", "purple", "seagreen4", "salmon2","orangered2")[as.factor(Environment)]

lda_echo_log2 <- lda(Environment ~ Fc + Fp, data = freq_log)
summary(lda_echo_log2)

# Calculating discriminant scores:

lscore_echo_log2 <- cbind(Fc, Fp)%*%lda_echo_log2$scaling 

#Plotting each measurement score against the Species variable 
plot(x = lscore_echo_log2[,1],lscore_echo_log2[,2],
     col = alpha(environment_lda_log2, 0.5),
     pch = 19, 
     cex = 1.5,
     xlab = "1st canonical variate",
     ylab = "2nd canonical variate",
     asp = 1,
     main = "Discriminant analysis")

legend("topright", 1, pch = 19, 
       legend = c("Riverine", "Coastal","Oceanic", "Polar", "Mixed")
       [unique(as.factor(Environment))], col = unique(environment_lda_log2))

```

## LDA or Canonical Variate Analysis -- model 3 (frequency, peak to peak source level and bandwidth)

```{r}

attach(echo_log_all1)

# Setting up color schemes
environment_lda_log3 <- c("turquoise3", "purple", "seagreen4", "salmon2","orangered2")[as.factor(Environment)]

lda_echo_log3 <- lda(Environment ~ SLpp + SLRMS + Fc + Fp + Duration + QRMS, data = echo_log_all)
lda_echo_log3

# Calculating discriminant scores:

lscore_echo_log3 <- cbind(SLpp, SLRMS, Fc, Fp, Duration, QRMS)%*%lda_echo_log3$scaling 

#Plotting each measurement score against the Species variable 
plot(x = lscore_echo_log3[,1],lscore_echo_log3[,2],
     col = alpha(environment_lda_log3, 0.5),
     pch = 19, 
     cex = 1.5,
     xlab = "1st canonical variate",
     ylab = "2nd canonical variate",
     asp = 1,
     main = "Discriminant analysis")

legend("topleft", 1, pch = 19, legend = c("Riverine", "Coastal","Oceanic", "Polar", "Mixed")
       [unique(as.factor(Environment))], col = unique(environment_lda_log3))

```

```